Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2020-09-30 ms/main.tex
small changes and added TODOs during talking with DC

With both vertical and oblique transmission, $0<v<1$, the recursion system is more complex
and we focus on local rather than on global stability.
To proceed,  note that 
Eq.\ \ref{eq:horizontal} gives $\hat{p}'$ as a function of both $p'$ and $\tilde{p}'$.
%Equation 4 does not give $\hat{p}$---MF
Eq.\ \ref{eq:nextgen_adults_slimpify} gives $p'$ as a function of $\tilde{p}$, since $\hat{p}$ is given in Eq.\ \ref{eq:horizontal} as a function of $\tilde{p}$ and 
Eq.\ \ref{eq:nextgen_parents_simplified} gives $\tilde{p}'$ as a function of $\hat{p}$. 
Combining these equations, we find an equation for $\hat{p}'$ as a function of $\hat{p}$ (shown in Appendix~\autoref{sec:appendixA}),
determine the equilibria, namely, solutions of $\hat{p}' = \hat{p}$, and analyse their local stability.
%: an equilibrium $\hat{p}^*$ is locally stable when the derivative of $f(\hat{p})=\bar{w}(\hat{p}'-\hat{p})$ at the equilibrium is negative, $f'(\hat{p}^*)<0$.

@@

%Note that the right-hand side equals $\hat\gamma_1$.
This condition merges the conditions for fixation of cooperation and for stable coexistence, which both lead to the instability of $\hat{p}=0$, the fixation of defection.
Importantly, increasing social association $\alpha$ increases the boundary cost ($\partial \hat\gamma_1 / \partial \alpha > 0$), making it easier for cooperation to increase from rarity.
Similarly, increasing the horizontal transmission of cooperation increases the boundary ($\partial \hat\gamma_1 / \partial T_A > 0$), making it easier for cooperation to evolve.
However, increasing the horizontal transmission of defection can either increase or decrease the boundary, making it easier or harder for cooperation to evolve, respectively, depending if $T_A$ is high or low (i.e. if $T_A>1/(1+bv\alpha)$ or below).
%($\text{sign}\big(\partial \hat\gamma_1 / \partial T_B \big)= \text{sign}\big(b v \alpha T_A - (1-T_A)\big)$).

@@

In Corollary~\autoref{corollary:intermediate_association}, we examined the effect of social association on evolution of cooperation in the case of perfect vertical transmission ($v=1$).
A more general case can be stated for $0 < v \leq 1$.
Consider the social association boundaries 
\begin{equation}\begin{aligned}\label{eq:boundries_assortative_meeting_general_case}
  \hat{a}_1 = \frac{c\cdot v(1-T_A) -(T_A-T_B)(1+b-c)}{b\cdot v \cdot T_B}, \quad
  \hat{a}_2 = \frac{c\cdot v(1-T_B)-(T_A-T_B)}{b\cdot v\cdot T_A} \;.
\end{aligned}\end{equation}
Note that when $T_A>T_B$ then $\partial \hat a_2 / \partial v = b T_A (T_A-T_B) > 0$, and when $T_A<T_B$ then $\partial \hat a_1 / \partial v = b(1+b-c)T_B(T_A-T_B) < 0$.
Similarly, $\partial \hat a_1 / \partial T_A < 0$, $\partial \hat a_2 / \partial T_A < 0$, $\partial \hat a_2 / \partial T_B > 0$, and $\text{sign}(\partial \hat a_1 / \partial T_B) = \text{sign}\big(T_A (1+b-c) - (1-T_A)cv\big)$.

@@

\begin{corollary}[Intermediate association of transmission and cooperation]
\label{corollary:intermediate_association_res3}
Cooperation will increase from rarity if social association is high enough, specifically if
  \begin{equation} \label{eq:unequal_transmission_from_rarity_in_alpha_terms}
    \begin{aligned}
    T_A > T_B &\;\; \text{and} \;\; \hat a_2 < \alpha, \quad \text{or} \\
    T_A < T_B &\;\; \text{and} \;\; \hat a_1 < \alpha \,.
    \end{aligned}
    \end{equation} 
\end{corollary}
\autoref{fig:result3}cd illustrates this conditions.

Note that increasing the horizontal transmission of cooperation, $T_A$, decreases both boundaries ($\partial \hat a_1 / \partial T_A, \partial \hat a_2 / \partial T_A < 0$), thus facilitating evolution of cooperation. 
Increasing the horizontal transmission of defection, increases one boundary, ($\partial \hat a_2 / \partial T_B > 0$).
So, if there is horizontal bias for cooperation ($T_A>T_B$), decreasing that bias will make it harder for cooperation to evolve.
However, if there is horizontal bias for defection ($T_A<T_B$) then the effect of further increasing the horizontal transmission of defection, $T_B$, can either increase or decrease the threshold $\hat\alpha_1$, making it harder or easier for cooperation to evolve, respectively, depending if horizontal transmission of cooperation ($T_A$) is high or low (i.e. if $T_A > cv/(1+b-c+cv)$ or below).

Interestingly, when there is horizontal transmission bias for cooperation ($T_A>T_B$), then vertical transmission hinders the evolution of cooperation ($\partial \hat a_2 / \partial v > 0$). In contrast, when there is horizontal transmission bias for defection ($T_A<T_B$), vertical transmission facilitates the evolution of cooperation ($\partial \hat a_1 / \partial v < 0$).
\\

@@

Social association may also emerge from a population's structure.
Consider a  population colonizing a 2D grid of size 100-by-100, where each site is inhabited by one individual, similarly to \citet{lewin2020rockpaperscissors}.
Each individual is characterized by its phenotype which is either cooperator, $A$, or defector $B$.
In each generation individuals interact with their neighbors, i.e. individuals that inhabit adjacent sites, in a prisoner's dilemma game (\autoref{table:prisoner_payoff}) and with horizontal cultural transmission.
As before, $T_A$ and $T_B$ are the probabilities of transmitting phenotypes $A$ and $B$ during horizontal transmission.
At each generation, each individual is chosen to "initiate" interactions with probability 1/2.\ %"with probability 1/2" not clear - MWF
Initiators then interact with a random neighbor in a prisoners' dilemma game and a random neighbor (with replacement) for horizontal cultural transmission (as both the game and the transmission are symmetrical, the expected number of each of these interactions per individual per generation is 1).
Therefore, the effective social association $\alpha$ in this model is the probability that the same neighbor is picked for both interactions, or $\alpha=1/m$, where $m$ is the number of neighbors. On an infinite grid, $m=8$, but on a finite grid $m$ can be lower in edge neighborhoods close to the grid border.
%How are A, B initialized on the grid? – MWF

@@

Let $f(p)=\lambda(p'-p)$, where $\lambda>0$, and assume $f(p^*)=0$; i.e., $p^*$ is an equilibrium.
%HOW DO YOU KNOW $p^*\in(0,1)$?--MF
We want a condition for $|p'-p^*| < |p-p^*|$.

@@

    \caption{\textbf{Evolution of cooperation under vertical and horizontal cultural transmission.} 
The figure shows the global fixation of cooperation (green), global fixation of defection (red), fixation of either cooperation or defection depending on the initial conditions, i.e. unstable coexistence (blue), and stable coexistence of cooperation and defection (yellow).
In all cases the horizontal bias ($T_A-T_B$) is on the x-axis.
(\textbf{a-b})~The cost of cooperation $c$ is on the y-axis; the cost boundaries $\gamma_1$ and $\gamma_2$ (\autoref{eq:cost_boundaries}) are the solid and dashed lines, respectively. 
(\textbf{c})~social association $\alpha$ is on the y-axis; the social association boundaries $a_1$ and $a_2$ (\autoref{eq:boundries_assortative_meeting}) are the solid and dashed lines, respectively.
Here, $b=1.3$, $T_A=0.4$. (\textbf{a})~$\alpha = 0$.(\textbf{b})~$\alpha = 0.7$. (\textbf{c})~$c = 0.35$.
  	}
    \label{fig:result2}
\end{figure}

>>>>

With both vertical and oblique transmission, $0<v<1$, the recursion system is more complex
and we focus on local rather than on global stability.
To proceed,  note that 
Eq.\ \ref{eq:horizontal} gives $\hat{p}'$ as a function of both $p'$ and $\tilde{p}'$.
%TODO Equation 4 does not give $\hat{p}$---MWF
Eq.\ \ref{eq:nextgen_adults_slimpify} gives $p'$ as a function of $\tilde{p}$, since $\hat{p}$ is given in Eq.\ \ref{eq:horizontal} as a function of $\tilde{p}$ and 
Eq.\ \ref{eq:nextgen_parents_simplified} gives $\tilde{p}'$ as a function of $\hat{p}$. 
Combining these equations, we find an equation for $\hat{p}'$ as a function of $\hat{p}$ (shown in Appendix~\autoref{sec:appendixA}),
determine the equilibria, namely, solutions of $\hat{p}' = \hat{p}$, and analyse their local stability.
%: an equilibrium $\hat{p}^*$ is locally stable when the derivative of $f(\hat{p})=\bar{w}(\hat{p}'-\hat{p})$ at the equilibrium is negative, $f'(\hat{p}^*)<0$.

@@

%Note that the right-hand side equals $\hat\gamma_1$.
This condition merges the conditions for fixation of cooperation and for stable coexistence, which both lead to the instability of $\hat{p}=0$, the fixation of defection.
Importantly, increasing social association $\alpha$ increases the boundary cost ($\partial \hat\gamma_1 / \partial \alpha > 0$), making it easier for cooperation to increase from rarity.
Similarly, increasing the horizontal transmission of cooperation increases the boundary ($\partial \hat\gamma_1 / \partial T_A > 0$), making it easier for cooperation to evolve.
However, increasing the horizontal transmission of defection can either increase or decrease the boundary, making it easier or harder for cooperation to evolve, respectively, depending if $T_A$ is high or low (i.e. if $T_A>1/(1+bv\alpha)$ or below).
%($\text{sign}\big(\partial \hat\gamma_1 / \partial T_B \big)= \text{sign}\big(b v \alpha T_A - (1-T_A)\big)$).
%TODO talk about partial gamma 1 / partial v = -(TA-TB)(1-TB)

@@

In Corollary~\autoref{corollary:intermediate_association}, we examined the effect of social association on evolution of cooperation in the case of perfect vertical transmission ($v=1$).
A more general case can be stated for $0 < v \leq 1$.
Consider the social association boundaries 
\begin{equation}\begin{aligned}\label{eq:boundries_assortative_meeting_general_case}
  \hat{a}_1 = \frac{c\cdot v(1-T_A) -(T_A-T_B)(1+b-c)}{b\cdot v \cdot T_B}, \quad
  \hat{a}_2 = \frac{c\cdot v(1-T_B)-(T_A-T_B)}{b\cdot v\cdot T_A} \;.
\end{aligned}\end{equation}
Note that when $T_A>T_B$ then $\partial \hat a_2 / \partial v = b T_A (T_A-T_B) > 0$, and when $T_A<T_B$ then $\partial \hat a_1 / \partial v = b(1+b-c)T_B(T_A-T_B) < 0$.
Similarly, $\partial \hat a_1 / \partial T_A < 0$, $\partial \hat a_2 / \partial T_A < 0$, $\partial \hat a_2 / \partial T_B > 0$, and $\text{sign}(\partial \hat a_1 / \partial T_B) = \text{sign}\big(T_A (1+b-c) - (1-T_A)cv\big)$.
%TODO remove derivatives of a1

@@

\begin{corollary}[Intermediate association of transmission and cooperation]
\label{corollary:intermediate_association_res3}
Cooperation will increase from rarity if social association is high enough, specifically if
  \begin{equation} \label{eq:unequal_transmission_from_rarity_in_alpha_terms}
    \begin{aligned}
    T_A > T_B &\;\; \text{and} \;\; \hat a_2 < \alpha, \quad \text{or} \\
    T_A < T_B &\;\; \text{and} \;\; \hat a_1 < \alpha \,.
    \end{aligned}
    \end{equation} 
\end{corollary}
\autoref{fig:result3}cd illustrates this conditions.
%TODO fix for increasing from rarity, only using a2

Note that increasing the horizontal transmission of cooperation, $T_A$, decreases both boundaries ($\partial \hat a_1 / \partial T_A, \partial \hat a_2 / \partial T_A < 0$), thus facilitating evolution of cooperation. 
Increasing the horizontal transmission of defection, increases one boundary, ($\partial \hat a_2 / \partial T_B > 0$).
So, if there is horizontal bias for cooperation ($T_A>T_B$), decreasing that bias will make it harder for cooperation to evolve.

Interestingly, when there is horizontal transmission bias for cooperation ($T_A>T_B$), then vertical transmission hinders the evolution of cooperation ($\partial \hat a_2 / \partial v > 0$). In contrast, when there is horizontal transmission bias for defection ($T_A<T_B$), vertical transmission facilitates the evolution of cooperation ($\partial \hat a_1 / \partial v < 0$).
\\
%TODO switch corollary 7 and 6

@@

Social association may also emerge from a population's structure.
Consider a  population colonizing a 2D grid of size 100-by-100, where each site is inhabited by one individual, similarly to \citet{lewin2020rockpaperscissors}.
Each individual is characterized by its phenotype which is either cooperator, $A$, or defector $B$.
In each generation individuals interact with their neighbors, i.e. individuals that inhabit adjacent sites, in a prisoner's dilemma game (\autoref{table:prisoner_payoff}) and with horizontal cultural transmission.
As before, $T_A$ and $T_B$ are the probabilities of transmitting phenotypes $A$ and $B$ during horizontal transmission.
At each generation, each individual is chosen to "initiate" interactions with probability 1/2.\ %TODO "with probability 1/2" not clear - MWF
Initiators then interact with a random neighbor in a prisoners' dilemma game and a random neighbor (with replacement) for horizontal cultural transmission (as both the game and the transmission are symmetrical, the expected number of each of these interactions per individual per generation is 1).
Therefore, the effective social association $\alpha$ in this model is the probability that the same neighbor is picked for both interactions, or $\alpha=1/m$, where $m$ is the number of neighbors. On an infinite grid, $m=8$, but on a finite grid $m$ can be lower in edge neighborhoods close to the grid border.
%TODO How are A, B initialized on the grid? – MWF

@@

Let $f(p)=\lambda(p'-p)$, where $\lambda>0$, and assume $f(p^*)=0$; i.e., $p^*$ is an equilibrium.
%TODO HOW DO YOU KNOW $p^*\in(0,1)$?--MWF
We want a condition for $|p'-p^*| < |p-p^*|$.

@@

    \caption{\textbf{Evolution of cooperation under vertical and horizontal cultural transmission.} 
The figure shows the global fixation of cooperation (green), global fixation of defection (red), fixation of either cooperation or defection depending on the initial conditions, i.e. unstable coexistence (blue), and stable coexistence of cooperation and defection (yellow).
In all cases the horizontal bias ($T_A-T_B$) is on the x-axis.
(\textbf{a-b})~The cost of cooperation $c$ is on the y-axis; the cost boundaries $\gamma_1$ and $\gamma_2$ (\autoref{eq:cost_boundaries}) are the solid and dashed lines, respectively. 
(\textbf{c})~social association $\alpha$ is on the y-axis; the social association boundaries $a_1$ and $a_2$ (\autoref{eq:boundries_assortative_meeting}) are the solid and dashed lines, respectively.
Here, $b=1.3$, $T_A=0.4$. (\textbf{a})~$\alpha = 0$.(\textbf{b})~$\alpha = 0.7$. (\textbf{c})~$c = 0.35$.
  	}
    \label{fig:result2}
\end{figure}
%TODO add panel d?
